Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-05-27 Multiresolution_framework/ltx/manuscript.tex
minor edits to introduction just typos

\section{Introduction}
The human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing range from ion channels, to single neurons, to networks of millions of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE), to address this challenge.

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  Electroencephalography (EEG) and Magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also be descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (\vik{pulse to wave conversion}) and the activation function (\vik{wave to pulse conversion}) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials \vik{in populations of microscopic single neural models, such as integrate and fire neurons,} or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter, can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) \vik{as} mean firing rates of neural masses. The key factor linking the functions of \vik{neural mass models} to particular spatial scales is \vik{the connectivity}, hence the ability to infer a multi-resolution connectivity kernel will enable the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. \vik{Large-scale neuroinformatic efforts such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}) are on the way to enable patient-specific full brain modeling.} This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

% At each spatial scale indexed by $j$ (level of decomposition), the neural field can be written as
% \begin{equation}
% 	v_{j,t}\left(\mathbf{r}\right) = \begin{cases}
% 	\sum_{l \in \mathbb{Z}}x_{j,l}(t) \underbrace{\phi_{j,l}\left(\mathbf{r}\right)}_{\text{scaling function}} + \check{x}_{t,j,l}\underbrace{\psi_{j,l}\left(\mathbf{r}\right)}_{\text{mother wavelet}} & \text{if $j=0$},\\
% 	\sum_{l \in \mathbb{Z}}\check{x}_{t,j,l}\psi_{j,l}\left(\mathbf{r}\right) & \text{otherwise}
% \end{cases}
% \end{equation}
The dynamic neural field is decomposed using
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l} \underbrace{\phi_{j_0,l}\left(r\right)}_{\text{scaling functions}} + \sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l} \underbrace{\psi_{j,l}\left(r\right)}_{\text{wavelets}},
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the dynamic coefficients of the expansion, constituting the state vector at time $t$. The decomposition enables a separation of the static spatial characteristics, which are considered static functions, that are weighted by dynamic coefficients.

>>>>

\section{Introduction}
The human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing range from ion channels, to single neurons, to networks of millions of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE) to address this challenge.

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  electroencephalography (EEG) and magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also be descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (\vik{pulse to wave conversion}) and the activation function (\vik{wave to pulse conversion}) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials \vik{in populations of microscopic single neural models, such as integrate and fire neurons,} or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) \vik{as} mean firing rates of neural masses. The key factor linking the functions of \vik{neural mass models} to particular spatial scales is \vik{the connectivity}, hence the ability to infer a multi-resolution connectivity kernel will enable the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. \vik{Large-scale neuroinformatic efforts, such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling.} This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

% At each spatial scale indexed by $j$ (level of decomposition), the neural field can be written as
% \begin{equation}
% 	v_{j,t}\left(\mathbf{r}\right) = \begin{cases}
% 	\sum_{l \in \mathbb{Z}}x_{j,l}(t) \underbrace{\phi_{j,l}\left(\mathbf{r}\right)}_{\text{scaling function}} + \check{x}_{t,j,l}\underbrace{\psi_{j,l}\left(\mathbf{r}\right)}_{\text{mother wavelet}} & \text{if $j=0$},\\
% 	\sum_{l \in \mathbb{Z}}\check{x}_{t,j,l}\psi_{j,l}\left(\mathbf{r}\right) & \text{otherwise}
% \end{cases}
% \end{equation}
The dynamic neural field is decomposed using
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l} \underbrace{\phi_{j_0,l}\left(r\right)}_{\text{scaling functions}} + \sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l} \underbrace{\psi_{j,l}\left(r\right)}_{\text{wavelets}},
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the dynamic coefficients of the expansion, constituting the state vector at time $t$. The decomposition enables a separation of the spatial characteristics, which are considered static functions that are weighted by dynamic coefficients.
